---
title: "National Living Wage: Causal Impact of Age‑Eligibility Thresholds"
author: "Group X"
output: html_document
---

```{r setup, message = FALSE, warning = FALSE}
# ---- Packages ----
packages <- c(
  "tidyverse",   # data management & ggplot2
  "lubridate",   # dates
  "fixest",      # DiD / fixed‑effects
  "rdrobust",    # sharp RDD
  "patchwork"    # combine plots
)

invisible(lapply(packages, function(p) if (!requireNamespace(p, quietly = TRUE)) install.packages(p)))
lapply(packages, library, character.only = TRUE)

# ---- Parameters ----
data_dir   <- "data_rest"          # folder containing APS .dta files
survey_yrs <- 14:24                # APS years 2014‑2024

# helper to define weight var present in each year
weight_vars <- c("PWTA14", "PWTA16", "PWTA18", "PWTA20", "PWTA22")

# ---- Variable map ----
var_map <- c(
  AGE      = "age",     AGEP      = "age",
  SEX      = "sex",
  ILODEFR  = "emp_status",
  REFDTE   = "ref_date",
  ETHGR11  = "ethnicity", ETHGR3   = "ethnicity", ETHUKEUL = "ethnicity",
  GOR      = "region",   CTRY9D    = "region"
)
```

## 1. Load & Tidy APS Waves

```{r load-clean}
clean_wave <- function(year) {
  file <- list.files(data_dir, pattern = paste0("apsp_jd", year, "_.*\\.dta$"), full.names = TRUE)
  if (length(file) == 0) {
    warning(paste("File for", year, "not found")); return(tibble())
  }

  df <- read_dta(file)
  
  # keep only variables we know about incl. weights
  vars_keep <- intersect(names(df), c(names(var_map), weight_vars))
  df <- df %>% select(all_of(vars_keep))

  # rename core vars
  df <- df %>% rename_with(~ var_map[.x], .cols = names(var_map)[names(var_map) %in% names(df)])

  # unify weight column
  wt_name <- intersect(weight_vars, names(df))[1]
  df <- df %>% rename(weight = {{wt_name}})

  # derive variables
  df %>% mutate(
    year        = 2000 + year,
    unemployed  = case_when(emp_status == 2 ~ 1L, emp_status %in% c(1, 3) ~ 0L),
    inactive    = case_when(emp_status == 3 ~ 1L, emp_status %in% c(1, 2) ~ 0L),
    age         = as.numeric(age),
    survey_date = dmy(ref_date)
  ) %>%
    # filter ages around policy windows (18‑65)
    filter(between(age, 18, 65), !is.na(unemployed), !is.na(weight), !is.na(survey_date))
}

aps <- map_dfr(survey_yrs, clean_wave)
# clean negative codes in numeric columns
aps <- aps %>% mutate(across(where(is.numeric), ~ ifelse(.x < 0, NA, .x)))

aps <- aps %>% mutate(ym = format(survey_date, "%Y-%m"))  # year-month factor

aps <- aps %>% filter(between(age, 19,30)) ## filter for 19-30 year olds

cat("Combined observations:", nrow(aps),"\n")
```

## 2. Policy Timeline & Treatment Flags

```{r policy-flags}
policy_timeline <- tribble(
  ~reform,      ~age_threshold, ~implementation,
  "NLW_2016",          25,      ymd("2016‑04‑01"),
  "NLW_2021",          23,      ymd("2021‑04‑01"),
  "NLW_2024",          21,      ymd("2024‑04‑01")
)

aps <- aps %>% mutate(
  post_2016 = as.integer(survey_date >= ymd("2016‑04‑01")),
  post_2021 = as.integer(survey_date >= ymd("2021‑04‑01")),
  post_2024 = as.integer(survey_date >= ymd("2024‑04‑01")),

  age_25p   = as.integer(age >= 25),
  age_23p   = as.integer(age >= 23),
  age_21p   = as.integer(age >= 21),

  treat_2016 = post_2016 * age_25p,
  treat_2021 = post_2021 * age_23p,
  treat_2024 = post_2024 * age_21p,

  age_run25 = age - 25,
  age_run23 = age - 23,
  age_run21 = age - 21,

  event_time_2016 = year - 2016,
  event_time_2021 = year - 2021,
  event_time_2024 = year - 2024
)
```
## 2.1 parallel trends testing

aps <- aps %>%
  mutate(
    cohort = case_when(
      between(age, 25, 30) ~ "25-30",
      between(age, 23, 24) ~ "23-24",
      between(age, 21, 22) ~ "21-22",
      between(age, 19, 20) ~ "19-20",
      TRUE ~ NA_character_
    ))
    
# Compute average outcome by cohort and year
avg_trends <- aps %>%
  filter(year < 2024) %>%  # only pre-treatment
  group_by(cohort, year) %>%
  summarise(mean_outcome = weighted.mean(unemployed, weight, na.rm = TRUE)) %>%
  ungroup()

# Plot
ggplot(avg_trends, aes(x = year, y = mean_outcome, color = cohort)) +
  geom_line(size = 1) +
  labs(
    title = "Parallel Trends Check (Pre-Treatment)",
    x = "Year", y = "Unemployment rate",
    color = "Age group"
  ) +
  scale_x_continuous(breaks = unique(avg_trends$year))+
  theme_minimal()


## 3. Exploratory Visualisations

### Survey Timing vs. Age

```{r viz-age-survey}
ggplot(aps, aes(x = survey_date, y = age)) +
  geom_hex(bins = 40) +
  geom_vline(data = policy_timeline, aes(xintercept = implementation), colour = "red", linetype = "dashed") +
  labs(title = "APS interviews by Age and Date", x = "Survey date", y = "Age")
```

### Unemployment Rate around Threshold

### Pre/post 2016

```{r viz-1}
plot_rdd <- function(data, cut_age, post_var, title) {
  ggplot(data, aes(x = age - cut_age, y = unemployed)) +
    stat_summary_bin(bins = 20, fun = mean) +
    geom_vline(xintercept = 0, colour = "red", linetype = "dotted") +
    facet_wrap(vars({{post_var}}), labeller = as_labeller(c("0"="Pre","1"="Post"))) +
    labs(title = title, x = "Age relative to threshold", y = "Unemployment rate")
}

plot_rdd(aps %>% filter(between(age,18, 30)), 25, post_2016, "Visualisation: 2016 threshold (Age 25)")
```

### Pre/post 2021

```{r viz-2}
plot_rdd <- function(data, cut_age, post_var, title) {
  ggplot(data, aes(x = age - cut_age, y = unemployed)) +
    stat_summary_bin(bins = 20, fun = mean) +
    geom_vline(xintercept = 0, colour = "red", linetype = "dotted") +
    facet_wrap(vars({{post_var}}), labeller = as_labeller(c("0"="Pre","1"="Post"))) +
    labs(title = title, x = "Age relative to threshold", y = "Unemployment rate")
}

plot_rdd(aps %>% filter(between(age, 18, 30)), 23, post_2021, "Visualisation: 2021 threshold (Age 23)")
```

### Pre/post 2021

```{r viz-3}
plot_rdd <- function(data, cut_age, post_var, title) {
  ggplot(data, aes(x = age - cut_age, y = unemployed)) +
    stat_summary_bin(bins = 20, fun = mean) +
    geom_vline(xintercept = 0, colour = "red", linetype = "dotted") +
    facet_wrap(vars({{post_var}}), labeller = as_labeller(c("0"="Pre","1"="Post"))) +
    labs(title = title, x = "Age relative to threshold", y = "Unemployment rate")
}

plot_rdd(aps %>% filter(between(age, 18, 30)), 21, post_2024, "Visualisation: 2024 threshold (Age 21)")
```

## 4. Difference‑in‑Differences

**4.1 DiD unemployment with basic controls**

```{r did-est}

# using the : to keep the DiD term only since age as IV is collinear with age_25p
did_2016 <- feols(unemployed ~ post_2016:age_25p |
                  age + ym, data = aps, weights = ~weight, cluster = ~ym
                  ) 
summary(did_2016)


# repeat for the main 2021 age reduction
did_2021 <- feols(unemployed ~ post_2021:age_23p |
                  age + ym, data = aps, weights = ~weight, cluster = ~ym
                  )
summary(did_2021)

# repeat for the 2024 further reduction
did_2024 <- feols(unemployed ~ post_2024:age_21p |
                  age + ym, data = aps, weights = ~weight, cluster = ~ym
                  )
summary(did_2024)

```

**4.2 DiD unemployment with extra controls (sex, ethnicity and region)**

No significant impact on age categories beyond 2016 introduction which suggests unemployment increased for NWL cohorts compared to age groups below the threshold.

#TODO: review this section, use other group closer to 25 year-olds as affected groups, review cluster applicability, etc.

```{r did-est-2}

did_2016 <- feols(unemployed ~ post_2016:age_25p |
                  age + ym + sex + ethnicity + region,
                  data = aps, weights = ~weight,
                  se = "twoway", cluster = ~ym + age
                  ) 
summary(did_2016)

# repeat for the main 2021 age reduction
did_2021 <- feols(unemployed ~ post_2021:age_23p |
                  age + ym + sex + ethnicity + region,
                  data = aps, weights = ~weight,
                  se = "twoway", cluster = ~ym + age
                  ) 
summary(did_2021)

# repeat for the 2024 further reduction
did_2024 <- feols(unemployed ~ post_2024:age_21p |
                  age + ym + sex + ethnicity + region,
                  data = aps, weights = ~weight,
                  se = "twoway", cluster = ~ym + age
                  ) 
summary(did_2024)

```

**4.3 DiD inactivity with basic controls**

If spec is correct, increase in NLW appears to have reduced economic inactivity for 2016 and 2021, 2024 events.

```{r did-est-3}

# using the : to keep the DiD term only since age as IV is collinear with age_25p
did_2016 <- feols(inactive ~ post_2016:age_25p |
                  age + ym, data = aps, weights = ~weight, cluster = ~ym
                  ) 
summary(did_2016)


# repeat for the main 2021 age reduction
did_2021 <- feols(inactive ~ post_2021:age_23p | 
                  age + ym, data = aps, weights = ~weight, cluster = ~ym
                  )
summary(did_2021)

# repeat for the 2024 further reduction
did_2024 <- feols(inactive ~ post_2024:age_21p | 
                  age + ym, data = aps, weights = ~weight, cluster = ~ym
                  )
summary(did_2024)
```

**4.4 DiD inactivity with extra controls**

Similar results with extra controls in place and two way clustered standard errors around year-month and age.

```{r did-est-4}

did_2016 <- feols(inactive ~ post_2016:age_25p |
                  age + ym + sex + ethnicity + region,
                  data = aps, weights = ~weight,
                  se = "twoway", cluster = ~ym + age
                  ) 
summary(did_2016)

# repeat for the main 2021 age reduction
did_2021 <- feols(inactive ~ post_2021:age_23p |
                  age + ym + sex + ethnicity + region,
                  data = aps, weights = ~weight,
                  se = "twoway", cluster = ~ym + age
                  ) 
summary(did_2021)

# repeat for the 2024 further reduction
did_2024 <- feols(inactive ~ post_2024:age_21p |
                  age + ym + sex + ethnicity + region,
                  data = aps, weights = ~weight,
                  se = "twoway", cluster = ~ym + age
                  ) 
summary(did_2024)

```

## 5. Event study Estimates (WiP)

(I think I did this wrong)

```{r event-est}

## Event-study around the 2016 cut-off
aps <- aps %>% 
  mutate(month_rel = floor(time_length(
    survey_date - as.Date("2016-04-01"), "month")
    )
  )

es_2016 <- feols(
  unemployed ~ i(month_rel, age_25p, ref = -1) | age + ym,
  data     = aps %>% filter(between(month_rel, -24, 24)),
  weights  = ~weight,
  cluster  = ~ym
)

iplot(es_2016, ref.line = 0,
      xlab = "Months relative to Apr-2016",
      main = "Event-study: unemployment gap 25+ vs <25")


```

## 6. Sharp RDD Estimates (WiP)

```{r rdd-est}
rdd_2016 <- rdrobust(y = aps$unemployed, x = aps$age_run25, c = 0,  masspoints = "adjust")
print(rdd_2016$coef)  # Treatment effect at age 25 in 2016 regime

```

Repeat for later reforms:

```{r rdd-loop, eval=FALSE}
rdd_res <- list(
  `2016` = rdrobust(aps$unemployed, aps$age_run25, c = 0, subset = aps$survey_date >= ymd("2015‑04‑01") & aps$survey_date < ymd("2017‑04‑01")),
  `2021` = rdrobust(aps$unemployed, aps$age_run23, c = 0, subset = aps$survey_date >= ymd("2020‑04‑01") & aps$survey_date < ymd("2022‑04‑01")),
  `2024` = rdrobust(aps$unemployed, aps$age_run21, c = 0, subset = aps$survey_date >= ymd("2023‑04‑01") & aps$survey_date < ymd("2025‑04‑01"))
)
map(rdd_res, ~ .x$coef)
```
