---
title: "National Living Wage: Causal Impact of Age‑Eligibility Thresholds"
author: "Group X"
output: html_document
---

```{r setup, message = FALSE, warning = FALSE}
# ---- Packages ----
packages <- c(
  "tidyverse",   # data management & ggplot2
  "lubridate",   # dates
  "fixest",      # DiD / fixed‑effects
  "haven"        # load .dta files
)

invisible(lapply(packages, function(p) if (!requireNamespace(p, quietly = TRUE)) install.packages(p)))
lapply(packages, library, character.only = TRUE)

# helper to define weight var present in each year
weight_vars <- c("PWTA14C", "PWTA16C", "PWTA17C", "PWTA18C", "PWTA20C", "PWTA22C")

# ---- Variable map ----
var_map <- c(
  AGE      = "age",     AGEP      = "age",
  SEX      = "sex",
  ILODEFR  = "emp_status",
  REFDTE   = "ref_date",
  ETHGR11  = "ethnicity", ETHGR3   = "ethnicity", ETHUKEUL = "ethnicity",
  HIQUL11D = "hi_qual", HIQUL15D = "hi_qual",
  GOR9d    = "region",   CTRY9D    = "region"
)
```

## 1. Load & Tidy APS Waves

```{r}

# ---- Setup ----
library(tidyverse)
library(fixest)
library(haven)

# ---- Load Data ----
lfs_js15raw <- read_dta("data_rest/five_q_longitudinal_js15-js16_eul_lgwt17.dta")
lfs_od15raw <- read_dta("data_rest/five_q_longitudinal_od15-od16_eul_lgwt17.dta")

# ---- Reshape to Long Format ----
reshape_lfs_long <- function(df, id_var = "PERSID") {
  quarters <- 1:5
  df_long <- map_dfr(quarters, function(q) {
    df %>% 
      transmute(
        PERSID = .data[[id_var]],
        quarter = q,
        age = .data[[paste0("AGE", q)]],
        emp_status = .data[[paste0("ILODEFR", q)]],
        weight = .data[[paste0("lgwt17")]]
      ) %>%
      filter(!is.na(age), !is.na(emp_status))
  })
  return(df_long)
}

lfs_js15 <- reshape_lfs_long(lfs_js15raw)
lfs_od15 <- reshape_lfs_long(lfs_od15raw)

# ---- Combine Panels ----
lfs_panel <- bind_rows(lfs_js15, lfs_od15) %>% 
  filter(between(age, 23, 27)) %>%
  mutate(
    treated = as.integer(age >= 25),
    post = as.integer(quarter >= 3),  # Treat Q3+ as post-reform (e.g., Apr-Jun 2016 onward)
    treat_post = treated * post,
    unemployed = case_when(
      emp_status == 2 ~ 1L,
      emp_status %in% c(1, 3) ~ 0L,
      TRUE ~ NA_integer_
    )
  )

# ---- Event Study: Create Relative Time ----
lfs_panel <- lfs_panel %>%
  group_by(PERSID) %>%
  mutate(rel_quarter = quarter - min(quarter)) %>%
  ungroup()

# ---- Estimate Event-Study Model ----
es_model <- feols(
  unemployed ~ i(rel_quarter, treated, ref = 1) | PERSID + quarter,
  data = lfs_panel,
  weights = ~weight,
  cluster = ~PERSID
)

# ---- Plot Coefficients ----
iplot(es_model, ref.line = 0,
      xlab = "Quarters relative to reform",
      main = "Event Study: NLW 2016 - Unemployment")

# ---- Wald Test for Pre-Trends ----
pre_vars <- grep("rel_quarter::[0-1]:treated", names(coef(es_model)), value = TRUE)
wald(es_model, pre_vars)


```


```{r load-clean}
# Function to process APS data
process_aps_data <- function(file_path) {
  df <- read_dta(file_path)
  
  # keep only variables we know about incl. weights
  vars_keep <- intersect(names(df), c(names(var_map), weight_vars))
  df <- df %>% select(all_of(vars_keep))
  
  # rename core vars - handle missing values safely
  if (exists("var_map") && length(var_map) > 0) {
    for (old_name in names(var_map)) {
      new_name <- var_map[[old_name]]
      if (!is.na(new_name) && old_name %in% names(df)) {
        names(df)[names(df) == old_name] <- new_name
      }
    }
  }
  
  # unify weight column
  if (exists("weight_vars") && length(weight_vars) > 0) {
    wt_candidates <- intersect(weight_vars, names(df))
    if (length(wt_candidates) > 0) {
      wt_name <- wt_candidates[1]
      names(df)[names(df) == wt_name] <- "weight"
    }
  }
  
  # derive variables with type checking
  df <- df %>% mutate(
    
    # Create employment status variables if emp_status exists
    unemployed = if ("emp_status" %in% names(df)) {
      case_when(
        emp_status == 2 ~ 1L, 
        emp_status %in% c(1, 3) ~ 0L,
        TRUE ~ NA_integer_
      )
    } else {
      NA_integer_
    },
    
    inactive = if ("emp_status" %in% names(df)) {
      case_when(
        emp_status == 3 ~ 1L, 
        emp_status %in% c(1, 2) ~ 0L,
        TRUE ~ NA_integer_
      )
    } else {
      NA_integer_
    },
    
    # Convert age to numeric if it exists
    age = if ("age" %in% names(df)) {
      as.numeric(age)
    } else {
      NA_real_
    },
    
    # Create survey_date if ref_date exists
    survey_date = if ("ref_date" %in% names(df)) {
      dmy(ref_date)
    } else {
      as.Date(NA)
    }
  )
  
  # filter ages around policy windows (18‑65)
  required_cols <- c("age", "unemployed", "weight", "survey_date")
  existing_cols <- intersect(required_cols, names(df))
  
  if (length(existing_cols) == length(required_cols)) {
    df <- df %>% 
      filter(
        between(age, 18, 65), 
        !is.na(unemployed), 
        !is.na(weight), 
        !is.na(survey_date)
      )
  } else {
    warning("Some required columns missing, skipping filtering: ", 
            paste(setdiff(required_cols, existing_cols), collapse = ", "))
  }
  
  # clean negative codes in numeric columns
  df <- df %>% mutate(across(where(is.numeric), ~ ifelse(.x < 0, NA, .x)))
  
  # create year-month factor if survey_date exists
  if ("survey_date" %in% names(df) && !all(is.na(df$survey_date))) {
    df <- df %>% mutate(ym = format(survey_date, "%Y-%m"))
  }
  
  return(df)
}

# Process both datasets
aps16 <- process_aps_data("data_rest/aps_3yr_jan15dec17_eul.dta")
aps21 <- process_aps_data("data_rest/aps_3yr_jan20dec22_eul.dta")

# Print summary information
cat("APS16 observations:", nrow(aps16), "\n")
cat("APS21 observations:", nrow(aps21), "\n")

# Print column names to verify processing
cat("APS16 columns:", paste(names(aps16), collapse = ", "), "\n")
cat("APS21 columns:", paste(names(aps21), collapse = ", "), "\n")
```

## 2. Policy Timeline & Treatment Flags

```{r policy-flags}
policy_timeline <- tribble(
  ~reform,      ~age_threshold, ~implementation,
  "NLW_2016",          25,      ymd("2016‑04‑01"),
  "NLW_2021",          23,      ymd("2021‑04‑01")
)

aps16 <- aps16 %>% mutate(
  post_2016 = as.integer(survey_date >= ymd("2016‑04‑01")),

  age_25p   = as.integer(age >= 25),

  treat_2016 = post_2016 * age_25p
)

aps21 <- aps21 %>% mutate(
  post_2021 = as.integer(survey_date >= ymd("2021‑04‑01")),

  age_23p   = as.integer(age >= 23),

  treat_2021 = post_2021 * age_23p
)
```

Variable presence in the dataset test, no need to run this chunk for further analysis

```{r}
#Test for variable presence

lfs_js15raw <- read_dta("data_rest/five_q_longitudinal_js15-js16_eul_lgwt17.dta")
lfs_od15raw <- read_dta("data_rest/five_q_longitudinal_od15-od16_eul_lgwt17.dta")

library(stringr)
check_weight_stringr <- function(df) {
  # Check column names
  weight_names <- str_detect(names(df), regex("year", ignore_case = TRUE))
  
  # Check labels
  weight_labels <- c()
  for (col in names(df)) {
    label <- attr(df[[col]], "label")
    if (!is.null(label) && str_detect(label, regex("year", ignore_case = TRUE))) {
      weight_labels <- c(weight_labels, col)
    }
  }
  
  return(list(
    names = names(df)[weight_names],
    labels = weight_labels
  ))
}

check_weight_stringr(lfs_js15raw)
check_weight_stringr(lfs_od15raw)


get_overlapping_persids <- function(df1, df2) {
  # Find common PERSID values
  common_ids <- intersect(df1$PERSID, df2$PERSID)
  
  cat("Number of overlapping PERSID values:", length(common_ids), "\n")
  
  if (length(common_ids) > 0) {
    cat("First 10 overlapping PERSID values:\n")
    print(head(common_ids, 10))
  } else {
    cat("No overlapping PERSID values found.\n")
  }
  
  return(common_ids)
}

get_overlapping_persids(lfs_js15raw, lfs_od15raw)

```

## 3. Exploratory Visualisations

### Survey Timing vs. Age

```{r viz-age-survey}
ggplot(aps16, aes(x = survey_date, y = age)) +
  geom_hex(bins = 40) +
  geom_vline(data = policy_timeline, aes(xintercept = implementation), colour = "red", linetype = "dashed") +
  labs(title = "APS interviews by Age and Date", x = "Survey date", y = "Age")
```

### Unemployment Rate around Threshold

### Pre/post 2016

```{r viz-1}
plot_rdd <- function(data, cut_age, post_var, title) {
  ggplot(data, aes(x = age - cut_age, y = unemployed)) +
    stat_summary_bin(bins = 20, fun = mean) +
    geom_vline(xintercept = 0, colour = "red", linetype = "dotted") +
    facet_wrap(vars({{post_var}}), labeller = as_labeller(c("0"="Pre","1"="Post"))) +
    labs(title = title, x = "Age relative to threshold", y = "Unemployment rate")
}

plot_rdd(aps16 %>% filter(between(age,18, 30)), 25, post_2016, "Visualisation: 2016 threshold (Age 25)")
```

### Pre/post 2021

```{r viz-2}
plot_rdd <- function(data, cut_age, post_var, title) {
  ggplot(data, aes(x = age - cut_age, y = unemployed)) +
    stat_summary_bin(bins = 20, fun = mean) +
    geom_vline(xintercept = 0, colour = "red", linetype = "dotted") +
    facet_wrap(vars({{post_var}}), labeller = as_labeller(c("0"="Pre","1"="Post"))) +
    labs(title = title, x = "Age relative to threshold", y = "Unemployment rate")
}

plot_rdd(aps21 %>% filter(between(age, 18, 30)), 23, post_2021, "Visualisation: 2021 threshold (Age 23)")
```

## 4. Difference‑in‑Differences (revised to exclude already treated groups)

**4.1 DiD unemployment with basic controls**

```{r did-est}

# using the : to keep the DiD term only since age as IV is colinear with age_25p
did_2016 <- feols(unemployed ~ post_2016:age_25p |
                  age + ym, data = aps16,
                  weights = ~weight, cluster = ~ym
                  ) 
summary(did_2016)


# repeat for the main 2021 age reduction
did_2021 <- feols(unemployed ~ post_2021:age_23p |
                  age + ym, data = aps21 %>% filter(between(age, 18, 24)), # added filter to exclude older, already treated groups
                  weights = ~weight, cluster = ~ym
                  )
summary(did_2021)

```

**4.2 DiD unemployment with extra controls (sex, ethnicity, region and highest quali)**

No significant impact on age categories beyond 2016 introduction which suggests unemployment increased for NWL cohorts compared to age groups below the threshold.

#TODO: review this section, use other group closer to 25 year-olds as affected groups, review cluster applicability, etc.

```{r did-est-2}

did_2016 <- feols(unemployed ~ post_2016:age_25p |
                  age + ym + sex + ethnicity + region + hi_qual,
                  data = aps16 %>% filter(between(age, 23, 26)), weights = ~weight,
                  se = "twoway", cluster = ~ym + age
                  ) 
summary(did_2016)

# repeat for the main 2021 age reduction
did_2021 <- feols(unemployed ~ post_2021:age_23p |
                  age + ym + sex + ethnicity + region  + hi_qual,
                  data = aps21 %>% filter(between(age, 21, 24)), weights = ~weight,
                  se = "twoway", cluster = ~ym + age
                  ) 
summary(did_2021)


```


**4.3 DiD inactivity with basic controls**

If spec is correct, increase in NLW appears to have reduced economic inactivity for 2016 and 2021, 2024 events.

```{r did-est-3}

# using the : to keep the DiD term only since age as IV is collinear with age_25p
did_2016 <- feols(inactive ~ post_2016:age_25p |
                  age + ym, data = aps16, weights = ~weight, cluster = ~ym
                  ) 
summary(did_2016)


# repeat for the main 2021 age reduction
did_2021 <- feols(inactive ~ post_2021:age_23p | 
                  age + ym, data = aps21 %>% filter(between(age, 18, 24)),
                  weights = ~weight, cluster = ~ym
                  )
summary(did_2021)
```

**4.4 DiD inactivity with extra controls**

Similar results with extra controls in place and two way clustered standard errors around year-month and age.

```{r did-est-4}

did_2016 <- feols(inactive ~ post_2016:age_25p |
                  age + ym + sex + ethnicity + region  + hi_qual,
                  data = aps16 %>% filter(between(age, 23, 26)), weights = ~weight,
                  se = "twoway", cluster = ~ym + age
                  ) 
summary(did_2016)

# repeat for the main 2021 age reduction
did_2021 <- feols(inactive ~ post_2021:age_23p |
                  age + ym + sex + ethnicity + region  + hi_qual,
                  data = aps21 %>% filter(between(age, 21, 24)),
                  weights = ~weight,
                  se = "twoway", cluster = ~ym + age
                  ) 
summary(did_2021)


```

## 6. Event study Estimates (WiP)

```{r event-est}
## Event-study around the 2016 cut-off
aps16 <- aps16 %>% 
  mutate(month_rel_16 = floor(time_length(
    survey_date - as.Date("2016-04-01"), "month")
    )
  )

es_2016 <- feols(
  unemployed ~ i(month_rel_16, age_25p, ref = -1:-1)| 
    age + ym + sex + ethnicity + region  + hi_qual,
  data     = aps16 %>% 
    filter(between(month_rel_16, -12, 12)) %>% filter(between(age, 24, 25)),
  weights  = ~weight,
  se = "twoway", cluster = ~ym + age
)

iplot(es_2016, ref.line = 0,
      xlab = "Months relative to Apr-2016",
      main = "Event-study: unemployment gap 25+ vs <25")

## Event-study around the 2021 cut-off
aps21 <- aps21 %>% 
  mutate(month_rel_21 = floor(time_length(
    survey_date - as.Date("2021-04-01"), "month")
    )
  )

es_2021 <- feols(
  unemployed ~ i(month_rel_21, age_23p, ref = -1:-1)| 
    age + ym + sex + ethnicity + region  + hi_qual,
  data     = aps21 %>%
    filter(between(month_rel_21, -12, 12)) %>% filter(between(age, 22, 23)),
  weights  = ~weight,
  se = "twoway", cluster = ~ym + age
)

iplot(es_2021, ref.line = 0,
      xlab = "Months relative to Apr-2021",
      main = "Event-study: unemployment gap 23+ vs <23")
```

6.1 Wald test

```{r}

# Wald Test for 2016 event pre-trend
# Get the names of all "lead" coefficients (k < 0)
pre_vars_16 <- grep("^month_rel_16::-[0-9]+:age_25p$",
                    names(coef(es_2016)),
                    value = TRUE)

# Cluster-robust Wald test (inherits the vcov from es_2016)
wald_pre_16 <- wald(es_2016, pre_vars_16)

print(wald_pre_16)

# Wald Test for 2021 event pre-trend
# Get the names of all "lead" coefficients (k < 0)
pre_vars_21 <- grep("^month_rel_21::-[0-9]+:age_23p$",
                    names(coef(es_2021)),
                    value = TRUE)

# Cluster-robust Wald test
wald_pre_21 <- wald(es_2021, pre_vars_21)

print(wald_pre_21)

```

```{r}
library(Matrix)

# 1. Variance-covariance sub-matrix for the leads
pre_vars_21 <- grep("^month_rel_21::-[0-9]+:age_23p$", names(coef(es_2021)), value = TRUE)
V21 <- vcov(es_2021)[pre_vars_21, pre_vars_21]

# 2. Check the numeric rank
cat("rank =", rankMatrix(V21)[1], " out of ", ncol(V21), "\n")

# 3. Condition number (large => near singular)
cat("condition number =", kappa(V21), "\n")

```