---
title: "Youth Unemployment and National Living Wage policy in the UK"
output: html_notebook
---

Setup
```{r}
library(did)
library(fixest)
library(tidyverse)
library(haven)
library(dplyr)
library(ggplot2)
library(rdrobust)

## load data
data14 <- read_dta("data_rest/apsp_jd14_eul_pwta18.dta")
data15 <- read_dta("data_rest/apsp_jd15_eul_pwta18.dta")
data16 <- read_dta("data_rest/apsp_jd16_eul_pwta18.dta")
data17 <- read_dta("data_rest/apsp_jd17_eul_pwta18.dta")
data18 <- read_dta("data_rest/apsp_jd18_eul_pwta18.dta")
data19 <- read_dta("data_rest/apsp_jd19_eul_pwta18.dta")
data20 <- read_dta("data_rest/apsp_jd20_eul_pwta22.dta")
data21 <- read_dta("data_rest/apsp_jd21_eul_pwta22.dta")
data22 <- read_dta("data_rest/apsp_jd22_eul_pwta22.dta")
data23 <- read_dta("data_rest/apsp_jd23_eul_pwta22.dta")
data24 <- read_dta("data_rest/apsp_jd24_eul_pwta22.dta")

# Define read_table()# Define vector of years
years <- 14:24

# Define subsetting function
subset_APS <- function(df, year) {
  # Define desired variables and fallback names
  var_map <- c(
    AGE = "age",
    AGEP = "age",
    SEX = "sex",
    ILODEFR = "emp",
    ACTHR = "hours",
    GRSSWK = "pay",
    HOURPAY = "hourly_pay",
    HPAYW = "hourly_pay",
    REFDTE = "ref_date",
    ETHGR11 = "ethnicity",
    ETHGR3 = "ethnicity",
    ETHUKEUL = "ethnicity",
    PWTA20 = "weight",
    PWTA16 = "weight",
    PWTA14 = "weight",
    PWTA22 = "weight",
    PWTA18 = "weight"
  )
  
  # Keep only variables that exist in the dataset
  available_vars <- intersect(names(var_map), names(df))
  
  df %>%
    select(all_of(available_vars)) %>%
    rename_with(~ var_map[.x], .cols = everything()) %>%
    filter(age >= 19, age <= 25) %>%
    mutate(year = 2000 + year)
}


# Apply to all loaded datasets
aps_panel <- bind_rows(lapply(years, function(y) {
  df_name <- paste0("data", y)
  subset_APS(get(df_name), y)
}))

# clean data
aps_panel[] <- lapply(aps_panel, function(x) {
  if (is.numeric(x)) ifelse(x < 0, NA, x) else x
})
```

# sorting cohorts
aps_subset <- aps_panel %>% filter(age >= 19, age <= 24)

aps_subset <- aps_subset %>%
  mutate(
    treatment_group = case_when(
      age %in% 19:20 ~ "control",
      age %in% 23:24 ~ "treated_23",
      age %in% 21:22 ~ "treated_21",
      TRUE ~ NA_character_
    )
  )

```

Parallel Trends plot
```{r}
avg_trends <- aps_subset %>%
  group_by(year, treatment_group) %>%
  summarise(mean_outcome = mean(pay, na.rm = TRUE), .groups = "drop")

ggplot(avg_trends, aes(x = year, y = mean_outcome, color = treatment_group)) +
  geom_line(linewidth = 1) +
  geom_point() +
  labs(
    title = "Parallel Trends: Mean Pay by Group",
    x = "Year",
    y = "Mean Pay",
    color = "Group"
  ) +
  theme_minimal()
```

Policy Variables

```{r}



```

